---
title: "Practice with dplyr and ggplot2"
author: "Gwendolyn Espinoza"
date: "DATA 2401"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Fancy Histograms!

We'll start by making some histograms.

```{r}
#install.packages("dslabs")
library(dslabs) 
data(heights)
glimpse(heights)

heights.2 <- heights
```

This data is the heights of humans, divided by their biological sex.

Use ggplot to make a histogram of all of the heights:

```{r}
heights.2 %>% ggplot(aes(x = height)) + geom_histogram()

```

Change up the binwidth and see how the plots change. Try 1, 5, 10, and 20

```{r}
heights.2 %>% ggplot(aes(x = height)) + geom_histogram(binwidth = 1)

```

Smooth this out to an emperical density with `geom_density()`

```{r}
heights.2 %>% ggplot(aes(x = height)) + geom_density()

```

Use a new argument in the `aes()`, `group =` to split this density by sex

```{r}
heights.2 %>% ggplot(aes(x = height, group = sex)) + geom_density()

```

OR we can do it with `color` or `fill.` If you say you want to color by sex, R knows that you want a different curve for each of them.

```{r}
heights.2 %>% ggplot(aes(x = height, color = sex)) + geom_density()

```

If you've used fill, then there is now a slight issue that they are overlapped. We can fix this with alpha transparency!

```{r}
heights.2 %>% ggplot(aes(x = height, fill = sex)) + geom_density(alpha = .2)

```

Let's make some boxplots of the same information.

```{r}
heights.2 %>% ggplot(aes(x = height)) + geom_boxplot()
```

## Quantatitive summaries:  

Find the mean and median overall. 

```{r}
heights %>% select(height) %>% summarize(mean(height))
heights %>% select(height) %>% summarize(median(height))
```

Find the mean and median for both groups. 

```{r}
heights %>% select(sex, height) %>% group_by(sex) %>% summarize( mean(height),median(height))

```

How tall is the tallest woman? How short is the shortest man? 

```{R}
heights %>% select(sex, height) %>% group_by(sex) %>% summarize( max(height))


```

## Presidental Elections Data

```{r}
#install.packages("pscl")
library(pscl) # loads in the package that has this data. 
## You might need to install this...

# data for presidental elections
votedata <-  presidentialElections
glimpse(votedata)

```

Let's look at the democratic vote by state for 2000. We can't use `geom_bar` for a bar chart, since we have the category in one variable and the "height" of the bar in another. We need `geom_col()`

Make a bar graph of the democratic vote by state in 2000. 

```{r}
ggplot(votedata %>% filter(year == 2000)) + geom_col(aes(x = state, y = demVote ))


```

Well this looks awful. We have two options: swap the x and y or the more fun sounding... Coordinate flip! 

Use `coord_flip()` on the previous graph to make it better. 

```{r fig.height=7, fig.width=5}

ggplot(votedata %>% filter(year == 2000)) + geom_col(aes(x = state, y = demVote )) + coord_flip()


```

I don't love the squashed together coordinates, but it's a display window issue.

So. This is a helpful graph, but it would be *more* helpful if it was ordered. 
Use `x = reorder(x_variable, y_variable)` in `aes()` to order the x variable by the y variable

```{r}
ggplot(votedata %>% filter(year == 2000), aes(x = reorder(state, demVote), y = demVote)) + geom_col()

```

So, what if I want to see what the north and south states did different?

start with a facet_wrap using the `south` variable:

```{r}
ggplot(votedata %>% filter(year == 2000), aes(x = reorder(state, demVote), y = demVote)) + geom_col() + facet_wrap(~ south)

```

Okay, that's not great. Lets *color* by south instead.

```{r}
ggplot(votedata %>% filter(year == 2000), aes(x = reorder(state, demVote), y = demVote, color = south)) + geom_col() 
```

I'm a good data scientist, so I want my plot to have a name! and my axes to have lables! Use `labs` to add a title, subtitle, and x and y labels.

```{r, eval = F}
ggplot(votedata %>% filter(year == 2000), aes(x = reorder(state, demVote), y = demVote, color = south)) + geom_col() +
  labs(
    title = 'Data in votes',
    subtitle = 'Data from the South and North',
    x = 'State',
    y = 'Dem Votes'
  )

```

You can move the legend with `theme(legend.position = "bottom")`

```{r, eval = F}
ggplot(votedata %>% filter(year == 2000), aes(x = reorder(state, demVote), y = demVote, color = south)) + geom_col() +
  labs(
    title = 'Data in votes',
    subtitle = 'Data from the South and North',
    x = 'State',
    y = 'Dem Votes'
  )+ theme(legend.position = "bottom")


```

What else could we facet by? years! Let's filter to year in 2008 and 2016, then facet by years.

```{r}
ggplot(votedata %>% filter(year %in% c(2008, 2016)), aes(x = reorder(state, demVote), y = demVote, color = south)) + geom_col() +
  labs(
    title = 'Data in votes',
    subtitle = 'Data from the South and North',
    x = 'State',
    y = 'Dem Votes'
  )+ theme(legend.position = "bottom") + facet_wrap(~ year)


```

We need to know who won! We could add a vertical line at 50 for who got more, to indicate the majority of votes. Adding the layer `geom_hline()` adds a horizontal line. (What do you guess `geom_vline()` would do?)

```{r}
ggplot(votedata %>% filter(year == 2008), aes(x = reorder(state, demVote), y = demVote, color = south)) + geom_point() +geom_hline(yintercept = 50) +
  labs(
    title = '2008 Presidential Election',
    subtitle = 'Data from the South and North',
    x = 'State',
    y = 'Dem Votes'
  )+ theme(legend.position = "bottom") 

```

### Getting fancy with a map!

When using geom_polygon or geom_map, you will typically need two data frames:

-   one contains the coordinates of each polygon (positions)
-   the other the values associated with each polygon (values).

An id variable links the two together. 

Run the below code to get a map graph. 

```{r}
library(maps)
votedata$state <- tolower(votedata$state)  ## states need to be lowercase for linking

states_map <- map_data("state") ## this gives us the lat and long for each point of each state.
  
map_plot <-  ggplot(data =  votedata %>% filter(year == 2008), aes(map_id = state)) +
    geom_map(aes(fill = demVote), map = states_map) +
    expand_limits(x = states_map$long, y = states_map$lat)
map_plot  
```

```{r}
map_plot <-  ggplot(data =  votedata %>% filter(year == 2016), aes(map_id = state)) +
   geom_map(aes(fill = demVote), map = states_map)+
   expand_limits(x = states_map$long, y = states_map$lat)
map_plot
```

What if I want a map that shows which of the states are "south"? What do I change?

```{r}
map_plot <-  ggplot(data =  votedata %>% filter(year == 2008), aes(map_id = state)) +
    geom_map(aes(fill = south), map = states_map) +
    expand_limits(x = states_map$long, y = states_map$lat)
map_plot

```


### Some more dplyr practice

I want to know the average democratic vote for N vs S, by year.

First, find the average democratic votes for the north and the south, every year. You'll need to do a *double* `group_by()` here. You do it in one call of the function.  

```{r}
a <- votedata %>% select(demVote,south) %>% group_by(demVote) %>% group_by(south) %>% summarize(mean(demVote))
a
```

Then, let's plot that! Pipe the result of your group_by and summarize to ggplot and geom_line(), with year on the x axis and your summarized value on the y axis. Color by the `south` variable. 

```{r}
#votedata %>% ggplot(aes(x = year, y = a )) + geom_line()

```

## Layering plots!

Penguins!

```{r}
library(palmerpenguins)
palmerpenguins::penguins
glimpse(penguins)
penguins.2 <- palmerpenguins::penguins
```

We can use boxplots to visualize the distribution of weight (body_mass_g) within each species:

```{r}
penguins.2 %>% ggplot(aes(x = body_mass_g)) + geom_boxplot()

```

What if we also want the points? Layering!! Add a geom_point to your existing boxplot. geom_boxplot + geom_point!

```{r}

penguins.2 %>% ggplot(aes(x = body_mass_g, y = sex)) + geom_boxplot() + geom_point()
```

But, these are all stacked up... to actually see them, use "geom_jitter" instead of points

```{r}
penguins.2 %>% ggplot(aes(x = body_mass_g, y = sex)) + geom_boxplot() + geom_jitter()

```

How to get the boxplots on top? The layers are plotted in the order you give them, so change to geom_point + geom_boxplot. (You might want to change the alpha on the boxplot to be able to see the plots under them)

```{r}
penguins.2 %>% ggplot(aes(x = body_mass_g, y = sex)) + geom_jitter() + geom_boxplot()

```

Maybe let's try replacing the boxplot with a `geom_violin()`?

```{r}
penguins.2 %>% ggplot(aes(x = body_mass_g, y = sex)) + geom_violin()



```

## If time: More Practice with Penguins 

-   Count the number of male and female penguins for each species.
-   Find the average penguin mass for each island
-   Find the average bill length by sex
-   Make a plot of bill_length and bill_depth, but only for the female penguins.
-   Make a plot of flipper length and body mass, faceted by island and colored by species.
-   Plot the density of the flipper length, colored by sex
-   Plot the density of the body mass, colored by year

```{r}



```
